Inferring the existence and direction of causal associations in the face of measurement error

### 2016-12-31 docs/manuscript.Rmd
updated

We used the Steiger test to infer the direction of causality between DNA methylation levels and gene expression levels. To do this we obtained a list of `r nrow(qtl_orig)` gene expression - DNA methylation associations as reported in Shakhbazov et al [-@Shakhbazov2016]. These were filtered to be located on the same chromosome, have robost correlations after correcting for multiple testing, and to share a SNP that had a robust cis-acting effect on both the DNA methylation probe and the gene expression probe. Using the reported summary data we used the Steiger test of two independent correlations to identify which direction of causality was most likely for each association. We found that the causal direction commonly goes in both directions (`r ctab("shakhtab")`) but methylation levels were more likely to influence expression levels than vice versa (probability of expression causing methylation = `r shakhtest3$p.value` (95% CI `r shakhtest3$conf.int[1]` - `r shakhtest3$conf.int[2]`) amongst `r sum(shakhbazov$dir_p < 0.05)` associations with $p_{Steiger} < 0.05$).

@@

In the presence of measurement error in $x$ and $y$, however, the Steiger test will instead be assessing the inequality $\rho_{g, x_o} > \rho_{g, y_o}$. In order to assess how reliable the Steiger test is in the presence of measurement imprecision, we can evaluate the range of potential values of measurement error in $x$ and $y$ over which the empirical result from the Steiger test would return the wrong value.

>>>>

We used the Steiger test to infer the direction of causality between DNA methylation levels and gene expression levels. To do this we obtained a list of `r nrow(qtl_orig)` gene expression - DNA methylation associations as reported in Shakhbazov et al [-@Shakhbazov2016]. These were filtered to be located on the same chromosome, have robost correlations after correcting for multiple testing, and to share a SNP that had a robust cis-acting effect on both the DNA methylation probe and the gene expression probe. Using the reported summary data we used the Steiger test of two independent correlations to identify which direction of causality was most likely for each association. We found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`). We then went on to predict the causal directions of the associations for varying levels of possible measurement error. `r cfig("shakhplot")` shows that the qualitative result of which of DNA methylation or gene expression is more commonly the causal factor is very strongly determined by measurement error.

@@

In the presence of measurement error in $x$ and $y$, however, the Steiger test will instead be assessing the inequality $\rho_{g, x_o} > \rho_{g, y_o}$, which can be simplified:

$$
\begin{aligned}
\rho_{g, x_O} & > \rho_{g, y_O} \\
\rho_{g, x} \rho_{x, x_O} & > \rho_{g,y}\rho_{y,y_O}\\
\rho_{g, x} \rho_{x, x_O} & > \rho_{g,x}\rho_{x,y}\rho_{y,y_O}\\
\rho_{x, x_O} & > \rho_{x,y}\rho_{y,y_O}
\end{aligned}
$$

In order to assess how reliable the Steiger test is in the presence of measurement imprecision, we can evaluate the range of potential values of measurement error in $x$ and $y$ over which the empirical result from the Steiger test would return the wrong value.
